\documentclass[%
 preprint,
superscriptaddress,
%groupedaddress,
%unsortedaddress,
%runinaddress,
%frontmatterverbose, 
%preprint,
%preprintnumbers,
%nofootinbib,
%nobibnotes,
%bibnotes,
 amsmath,amssymb,
 aps,
 linenumbers,
%pra,
%prb,
%rmp,
%prstab,
%prstper,
%floatfix,
]{revtex4-2}

\usepackage{physics}
\usepackage{booktabs}
\usepackage{braket}
\usepackage{qcircuit}
\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage[hidelinks]{hyperref}% add hypertext capabilities
\usepackage{makecell}
\usepackage{tensor}
\usepackage{xcolor}
\usepackage{threeparttable}
\usepackage{chemformula}
\newcommand{\sherry}[1]{#1}
\newcommand{\yuyang}[1]{\textcolor{blue}}
\counterwithin{subsection}{section}                     % 让subsection随section重置
\renewcommand{\thesubsection}{\thesection.\arabic{subsection}} 
\makeatletter
\def\switch@array{}
\makeatother

\begin{document}


\title{AdsMind: A Physics-Grounded Multi-agent System for Self-Correcting Discovery of Adsorption Configurations on Heterogeneous Catalyst Surfaces}
%
\author{Zongmin Zhang}
\affiliation{Department of Computer Science and Engineering, Hong Kong University of Science and Technology, Kowloon, Hong Kong 999077, China}
\author{Yuyang Lou}
\affiliation{Department of Chemistry, Hong Kong University of Science and Technology, Kowloon, Hong Kong 999077, China}
\author{Bowen Zhang}
\affiliation{Department of Chemistry, Hong Kong University of Science and Technology, Kowloon, Hong Kong 999077, China}
\author{Junwu Chen}
\affiliation{Laboratory of Artificial Chemical Intelligence (LIAC), EPFL, Lausanne, Switzerland}
\author{Ryo Kuroki}
\affiliation{Laboratory of Artificial Chemical Intelligence (LIAC), EPFL, Lausanne, Switzerland}
\affiliation{Platform Laboratory for Science \& Technology, Asahi Kasei Corporation, Tokyo, Japan}
\author{Xuan Vu Nguyen}
\affiliation{Laboratory of Artificial Chemical Intelligence (LIAC), EPFL, Lausanne, Switzerland}
\author{Edvin Fako}
\email{edvin.fako@epfl.ch}
\affiliation{Laboratory of Artificial Chemical Intelligence (LIAC), EPFL, Lausanne, Switzerland}
\author{Lixue Cheng}
\email{lixuecheng@ust.hk}
\affiliation{Department of Chemistry, Hong Kong University of Science and Technology, Kowloon, Hong Kong 999077, China}
\author{Philippe Schwaller}
\email{philippe.schwaller@epfl.ch}
\affiliation{Laboratory of Artificial Chemical Intelligence (LIAC), EPFL, Lausanne, Switzerland}
\date{\today}



\begin{abstract}
Identifying the most stable adsorbate-surface configuration is critical in heterogeneous catalysis and electrochemical energy conversion modeling, yet it remains computationally expensive and expert-intensive using traditional ab initio methods. Although machine learning force fields (MLFFs) have made great improvements on the evaluation efficiency, the upstream search strategy to decide which configurations to evaluate remains unsolved. Existing LLM-based agents for this task usually operate in an open-loop, single-pass regime, making them incapable of correcting themselves with wrong initial guesses.
We introduce \textbf{AdsMind} (\textbf{Ads}orption with \textbf{M}achine \textbf{I}ntelligence \textbf{\&} Fee\textbf{d}back), the first closed-loop multi-agent framework that enables autonomous error correction through MLFF relaxation feedback. This is achieved via three mechanisms implemented in five modules: Chemical slip detection to diagnose planned-vs-realized binding mismatches, FORBID constraints to prevent retrying failed site families, and convergence-based early stopping. With systematic ablations run via four LLM backends, AdsMind achieves near-perfect search reliability, i.e., 100\% on CMU20, 98.8\% on OCD-GMAE62, for the two evaluated datasets. Compared with a literature approach, Adsorb Agent, AdsMind reduces cross-backend energy variance by 2.42x, and consumes 5-25x fewer MLFF relaxations than brute-force baselines.
DFT validation (VASP/PBE) on six representative CMU20 systems reveals that the open-loop baseline exhibits qualitative sign errors on structurally complex adsorbates, while AdsMind preserves 100\% sign-correctness with competitive quantitative agreement. Chemical Slip further provides a physically grounded evaluation signal for LLM chemical reasoning quality.  AdsMind establishes a foundation for autonomous chemistry workflows by delivering reliability, efficiency, and interpretability simultaneously.
\end{abstract}


%\keywords{Suggested keywords}%Use showkeys class option if keyword
                              %display desired
\maketitle 

\section{Introduction}
\input{sections/1_Introduction}

\section{Method}
\label{sec:methods}
\input{sections/2_Method}

\section{Results}
\input{sections/3_Results}


\input{sections/4_DiscussionConclusion}

\section*{Data Availability}
\input{sections/5_Data_Availability}




\section*{Acknowledgements}
We acknowledge financial support from the Swiss National Science Foundation (SNSF) through the National Centre of Competence in Research (NCCR) Catalysis (grant number 180544). We also thank the EPFL Scientific IT and Application Support team for their computational support. Z.Z. thanks the support from Hong Kong Special Administrative Region Government Scholarship Fund Scholarship

\textbf{AI Disclosure Statement.}
The authors used generative AI tools (including large language models such as GPT-4, Claude, and Grammarly) for \textbf{grammar polishing, language editing, and minor phrasing improvements} in preparing this manuscript. All scientific content, analysis, interpretations, and conclusions were solely produced by the authors. The authors reviewed and edited the output of these tools and take full responsibility for the content of the published article.


\section*{Competing Interests}

The authors declare no competing interests.


\bibliography{journal_abbrev,refs}

\end{document}
%
% ****** End of file apssamp.tex ******
