\section{Discussion}

\subsection{What AdsMind contributes, and what it does not}

The results of this study, including the deliberately adversarial baselines introduced above, point to a nuanced characterization of closed-loop LLM agents that differs from the message conveyed by much of the prior literature.
Prior work on LLM agents for chemistry has tended to emphasize the \textit{capability} axis: what new kinds of tasks can an LLM accomplish that prior methods could not?
Our evidence suggests that for the specific problem of finding low-energy adsorbate configurations, the capability axis is not where closed-loop LLM reasoning wins.
Under identical MACE physics, random sampling with 20 relaxations matches AdsMind on ten of twenty CMU cases and beats it on six; heuristic enumeration of $\sim$50 canonical sites beats AdsMind on six of fifteen locked CMU cases; Adsorb-Agent's multi-configuration search reaches lower energies on all 12 comparable GPT-5.4 cases, and on 9 of 11 comparable cases in the historical GPT-4o run.
These are not artifacts of a single LLM backend: the closed-loop reliability and backend-convergence results hold across four distinct LLM backends and two independent benchmarks (CMU and OCD-GMAE), while the completed CMU20 five-variant ablation and MACE-large checks show that representative AdsMind trajectories remain reliable but force-field sensitive.

What closed-loop LLM reasoning contributes is along a different axis: the combined property of being \textit{reliable, compute-efficient, and interpretable} simultaneously.
Each of these attributes alone has precedents. Brute-force controls can be reliable.
Heuristic enumeration is interpretable (canonical sites). Single-attempt LLM planning can be MACE-efficient but brittle.
What is new in AdsMind is the co-occurrence: 100\% Full-variant success across every case and backend in the completed CMU20 and OCD-GMAE ablation matrices, 320/320 successful CMU20 iterative-variant runs versus 73/80 successful CMU20 1-Shot runs, 5--25$\times$ less MACE compute per single-backend run than the brute-force alternatives, and physically grounded reasoning artifacts (slip events, forbid constraints, site-mismatch fingerprints) that a practitioner can inspect, contest, and incorporate into downstream workflows.
This combination is a prerequisite for autonomous chemistry, where silent failure or irreproducibility in any one of these dimensions would break the pipeline.



\subsection{Claims and evaluation}
%% this section is moved from Introduction, but need a summary in Introduction
%% need to be integrated with the subsection above

This paper establishes four empirical claims through 600 controlled ablation attempts across two datasets and four LLM backends, plus matched baseline controls.
We now use the completed 20-case CMU and 10-case OCD-GMAE ablation matrices as mechanism tests, and use additional controls---CMU20 random placement, MACE-large sensitivity, and Adsorb-Agent single-configuration stress testing---to define the operating envelope:

\begin{description}
    \item[C$_1$ (reliability):] AdsMind Full achieves 100\% search success across every backend and every case on both completed ablation matrices: 80/80 CMU20 Full runs and 40/40 OCD-GMAE Full runs. On CMU20, the four iterative variants (Full, $-$Slip, $-$Forbid, $-$Term) also succeed on 320/320 case$\times$backend runs, whereas the 1-Shot control succeeds on 73/80; the seven failures are natural dissociation/reaction outcomes, not API or calculator failures. On the matched CMU overlap, single-pass Adsorb-Agent runs with GPT-4o and GPT-5.4 succeed on 73--80\% of cases, and the single-configuration Adsorb-Agent stress test produces no selected configuration on 6 of 20 CMU cases.
    \item[C$_2$ (backend convergence):] The iterative closed-loop architecture reduces energy variance across heterogeneous LLM backends. On the completed CMU20 matrix, all 80 Full runs succeed, the mean four-backend range is 0.153~eV, and 12 of 20 cases agree within 0.05~eV; the success-conditioned 1-Shot range is 0.359~eV, giving a 2.3$\times$ reduction toward a shared physical solution. The equivalent one-shot-to-full reduction on OCD-GMAE is also 2.3$\times$.
    \item[C$_3$ (compute efficiency):] A single AdsMind Full run uses approximately 4 MACE relaxations per case, compared to 20 (random), 25--98 (heuristic), and $\sim$21 (Adsorb-Agent). In the CMU20 best-of-four-backends reporting view, AdsMind matches or beats random placement on 14 of 20 cases under a 0.01~eV tolerance; in the locked 15-case comparison, a single-backend AdsMind run matches or beats heuristic enumeration on 9 of 15 cases.
    \item[C$_4$ (interpretability):] Closed-loop reasoning produces structured artifacts---slip events, forbid constraints, site-mismatch fingerprints---that serve as physically grounded diagnostics for LLM chemical reasoning. We show that one-shot slip rates correlate with surface composition complexity across all four backends, establishing Chemical Slip as a standalone evaluation signal for LLM science applications.
\end{description}


\subsection{Chemical Slip as a standalone LLM evaluation signal}

Chemical Slip---the mismatch between an LLM-planned adsorption site and the site realized after physical relaxation---was introduced as a feedback signal for the closed-loop agent.
The slip statistics aggregated across the CMU 20-case one-shot benchmark and the 200-run OCD-GMAE rep50 one-shot study reveal a secondary value: slip frequency, stratified by surface type and adsorbate, serves as a physically grounded proxy for LLM chemical reasoning quality that is independent of any downstream task metric.
On the CMU benchmark (see the Chemical Slip diagnostic section in the SI), the one-shot slip rate on intermetallic surfaces (71--77\%) far exceeds the rate on monometallic surfaces (20--40\%) across all four backends, indicating that LLM chemical priors are less reliable on complex alloy compositions under this relaxation protocol.
In contrast to text-based chemistry benchmarks such as ChemBench~\cite{mirza2024chembench}, which test chemistry knowledge and reasoning in language form, Chemical Slip evaluates \textit{applied} reasoning---whether the model's structural prediction survives physical simulation---and complements existing LLM evaluation frameworks for scientific domains.

\subsection{Why brute-force sampling wins on depth, and when it matters}

The energy-depth advantage of random placement and heuristic enumeration on complex surfaces requires explicit mechanistic interpretation.
On simple adsorbates with few competitive binding sites (H on Cu$_3$Ag(111), OH on Pt(100)), AdsMind Full matches the non-LLM baselines within 0.01~eV; the PES has few distinct basins and iterative refinement from any initial guess converges to the same low-energy basin.
On complex adsorbates (CH$_2$CH$_2$OH, OCHCH$_3$, ONN(CH$_3$)$_2$), the PES has many local minima separated by non-trivial barriers; iterative refinement from a single trajectory becomes trapped in whichever basin the first relaxation descends into, regardless of whether a deeper basin exists elsewhere.
Broad initial sampling circumvents this trap by seeding multiple trajectories in different basins.

This observation matters practically because it delineates when closed-loop search is the right tool.
For high-throughput catalyst screening where thousands of candidate surfaces must be down-selected and abundant parallel compute is available, brute-force methods can be more wall-clock efficient (all relaxations run in parallel) and their failure modes are tolerable (failed cases can be re-queued), even though they consume $5$--$25\times$ more MACE relaxations per case than AdsMind.
For autonomous exploration of a single promising candidate where failure is costly and interpretability is required, closed-loop refinement adds value.
A two-stage pipeline---brute-force screening followed by closed-loop refinement on top candidates---would combine the throughput of the former with the reliability of the latter; the concrete architectural integration is sketched at the end of the Conclusion.

\subsection{MLFF as surrogate: scope and validation}

All primary benchmark and matched-baseline experiments in this study use MACE-MP-0 small as the physical backend.
The agent's value is orthogonal to MLFF accuracy: it optimizes search strategy, not the energy function itself.
Adsorption energy in this study is operationally defined within the fixed MACE-MP-0 small protocol: all ablation comparisons, Wilcoxon tests, and backend-convergence metrics are evaluated on energies produced by the same locked MACE-small calculator, so within-protocol rankings between configurations on the same surface are self-consistent by construction.
We do not claim that MACE-small rankings match DFT rankings in absolute terms; rather, the MACE-large run described below serves as a sensitivity boundary that quantifies how far absolute energies shift under a higher-fidelity force field, without being treated as ground-truth validation.

To characterize this sensitivity, we repeated the GPT-5.4 AdsMind Full variant on all 20 CMU cases with MACE-large (GPU, float64, dispersion enabled), and on a 10-case representative slice of OCD-GMAE rep50.
MACE-large shifts absolute AdsMind energies substantially: on CMU20 the mean absolute shift relative to MACE-small is 1.255~eV, or 0.834~eV when the anomalous case~20 is excluded.
The direction of the shift is case-dependent, with 15/20 CMU cases becoming more stable and 5/20 becoming less stable under MACE-large.
Case~20 is an anomaly rather than a clean failure: the summary retains a non-dissociated best structure at $-2.402$~eV, while later dissociated analyses reach lower unretained energies.
On the OCD-rep10 sensitivity set, MACE-large succeeds on 9/10 cases, with case~043 failing naturally through slip and dissociation.
Because Adsorb-Agent was not rerun with MACE-large, this experiment should be interpreted as a force-field sensitivity check on AdsMind trajectories, not as a second AdsMind--Adsorb-Agent head-to-head.
Representative DFT validation remains necessary before interpreting the MLFF-ranked structures as final adsorption energetics beyond the fixed surrogate protocol.

Known MLFF failure modes---dispersion-dominated binding, strongly correlated $f$-electron systems, strong charge-transfer interfaces---define the current scope boundary of the framework.
Extending to these regimes will require either higher-fidelity MLFFs or hybrid MLFF--DFT workflows, but does not require changes to the agent architecture.

\subsection{Limitations and negative results}

\paragraph{Energy depth on complex adsorbates.}
On cases with multiple competitive binding sites and complex adsorbates (16, 17, 18, 19), AdsMind Full consistently produces higher energies than the broader-sampling baselines that succeed on those cases.
A five-seed control experiment (the default GPT-5.4 run plus four independent reruns on cases 16, 17, 18) reduces but does not remove this depth gap on the comparable Adsorb-Agent cases: the best-of-seeds AdsMind result remains 0.90~eV higher on case~16 and 1.20~eV higher on case~17, while GPT-5.4 Adsorb-Agent produces no valid trajectory files on case~18.
The mechanism is the iterative refinement trajectory's tendency to stay within the initial energy basin regardless of seed.
We do not regard this as a defect of the architecture---it is inherent to any single-trajectory closed-loop refinement---but as a specification of the operating envelope in which closed-loop refinement adds value.

\paragraph{FORBID over-pruning.}
On several cases in both CMU and OCD-GMAE, removing the FORBID constraint \textit{improves} the final energy.
On GPT-5.4 the pairwise Wilcoxon test for $-$Forbid trends toward significance (BH $p = 0.084$) with negative rank-biserial correlation ($r = -0.79$), indicating that explicit negative constraints over-prune the search space on surfaces where a lower-energy site was initially rejected.
Adaptive or soft-constraint variants of the FORBID mechanism, in which the strength of a negative constraint decays with additional evidence, merit future investigation.

\paragraph{Statistical power.}
Although the original 15-case CMU statistical package and the 10-case OCD-GMAE benchmark are large enough for Friedman tests to reach significance across all four backends (CMU: $p = 1.4\times10^{-6}$ to $2.2\times10^{-3}$; OCD-GMAE: $p = 8.2\times10^{-5}$ to $6.9\times10^{-3}$), many pairwise Wilcoxon tests do not survive BH correction.
We interpret individual pairwise comparisons primarily through effect sizes and bootstrap confidence intervals, and rely on the \textit{consistent direction} of the main full-versus-single-shot effect across four distinct LLM backends as a form of conceptual replication.

\paragraph{Scope.}
The current implementation handles flat and stepped crystalline surfaces with a single adsorbate species.
Extensions to defective surfaces, nanoparticles, co-adsorption, and solvated interfaces are architecturally straightforward but require additional validation.

\paragraph{LLM reproducibility and platform sensitivity.}
LLM APIs are black-box services subject to unannounced model updates; at temperature 0.0, outputs may still vary across API versions.
We mitigate this by recording configured model identifiers, API transport variants, and frozen experiment settings, and by releasing structured result files and analysis scripts so that all reported numbers can be independently reproduced from saved summaries.
The four-backend design further ensures that conclusions do not depend on any single LLM provider.
Where platform or MLFF-environment changes are introduced, absolute energies should be treated as sensitivity data rather than pooled with the locked ablation matrix.
All within-backend ablation comparisons in the main text use fixed frozen configurations and a single platform/physics stack for the compared runs.

\subsection{The uncomfortable truth: configuration-space coverage dominates}
%% this section is moved from Introduction, but need a summary in Introduction
%% this subsection should be part of the limitations and negative results

A second line of evidence, from exhaustive control experiments we report below, shows that on complex intermetallic surfaces with multiple competitive binding sites, the dominant factor in finding low-energy configurations is often simply \textit{the number of distinct initial configurations sampled}.
On the completed CMU20 random-control comparison, random placement of N=20 adsorbate poses, relaxed under the same MACE-small protocol, finds materially lower energies ($>0.01$~eV) than the best-of-four-backends AdsMind Full result on 6 of 20 cases (by up to 2.07~eV), while AdsMind is lower on 4 cases and 10 cases tie within 0.01~eV.
Heuristic enumeration of all high-symmetry sites (25--98 per case) beats AdsMind Full by $>0.01$~eV on 6 of 15 cases and matches within 0.01~eV on 8 cases.
Adsorb-Agent with its $\sim$21 MACE relaxations per case reaches lower energies than the GPT-5.4 AdsMind run on all 12 comparable GPT-5.4 cases; in the historical GPT-4o comparison against AdsMind Full best-across-backends, it wins 9 of 11 cases using a 0.01~eV tie threshold (one tie and one AdsMind win).
These results make clear that broader sampling, when it succeeds, matches or exceeds any closed-loop refinement strategy at the cost of 5--25$\times$ more MACE compute; the random-control winners are interpreted as protocol-level energy comparisons until their top structures receive the same anomaly filtering used for the agentic runs.

The question, then, is not ``which system finds the absolute lowest energy?''---an LLM agent cannot realistically out-sample a brute-force enumerator with sufficient budget---but rather: \textit{what is the contribution of closed-loop LLM reasoning when we hold compute, physics, and data fixed?}


\section{Conclusion}

We have presented AdsMind, a closed-loop multi-agent framework that integrates LLM-driven planning with iterative MLFF relaxation feedback for autonomous adsorption configuration search.
Three mechanisms---Chemical Slip detection, FORBID constraint memory, and autonomous termination---enable the agent to detect its own reasoning errors, avoid repeating them, and stop when further exploration is unproductive.

Through 600 controlled ablation attempts across four LLM backends (Gemini, Grok-4, GPT-5.4, Claude) on two independent benchmarks (20-case CMU and 10-case OCD-GMAE), together with a separate 200-run OCD-GMAE one-shot slip diagnostic and head-to-head comparisons against two non-LLM controls (random placement, heuristic enumeration) and two Adsorb-Agent runs (GPT-4o and GPT-5.4) under identical MACE physics, we establish that AdsMind's primary value lies in the combination of (i) 100\% Full-variant search success across every case and backend in both completed ablation matrices, with CMU20 iterative variants succeeding on 320/320 runs and the 1-Shot control exposing seven natural dissociation/reaction failures; (ii) cross-backend energy convergence (2.3$\times$ reduction in backend spread on CMU20); (iii) per-run compute efficiency (5--25$\times$ fewer MACE relaxations than brute-force controls); and (iv) interpretable reasoning artifacts in the form of slip events and forbid constraints.
Additional completed controls extend the reliability and scope boundary: the completed CMU20 random comparison shows that broader random sampling remains energetically competitive, the Adsorb-Agent single-configuration stress test confirms that Adsorb-Agent's depth is largely a breadth effect, and MACE-large runs quantify force-field sensitivity rather than providing DFT validation.
We also report a substantive negative result: on complex adsorbates, brute-force sampling reaches lower energies than closed-loop refinement, a trade-off that arises from single-trajectory iteration rather than from insufficient compute, and one that motivates future hybrid architectures.
Chemical Slip, introduced here as a feedback signal, additionally serves as a physically grounded diagnostic for LLM chemical reasoning quality---a standalone methodological contribution that may extend beyond the application in this work.

The central message is that closed-loop LLM reasoning is not a replacement for brute-force configuration search on well-resourced problems; it is the foundation for a new class of \textit{autonomous chemistry workflows} where reliability, efficiency, and interpretability must be achieved simultaneously.
A natural next step is to combine AdsMind's closed-loop error recovery with broader initial site enumeration, yielding an agent that inherits the depth of brute-force methods and the reliability of closed-loop refinement.

Code, frozen experiment configurations, structured result files, baseline implementations, and comparison scripts are available at \url{https://github.com/AI4QC/AdsMind}.
