All agent-side benchmark inputs, frozen configurations, runner scripts, and curated structured outputs used in this study are available in the AdsMind GitHub repository at \url{https://github.com/AI4QC/AdsMind}. The repository contains paper-facing summary tables, backend-comparison artifacts, and LaTeX table exports used in the manuscript under \texttt{research/agent\_eval/} and \texttt{research/results/}, as well as frozen experiment configurations and runner scripts for reproducing the reported benchmark and ablation summaries. Per-run logs, trajectories, and structure artifacts are treated as regenerable execution products. The reported benchmark and ablation summaries can be reproduced from the archived manifests, frozen configurations, and runner scripts.

The primary calculator is loaded through \texttt{mace.calculators.mace\_mp} with the MACE-MP-0 small checkpoint~\cite{batatia2023foundation}, CPU execution, float32 precision, and dispersion disabled. The MACE-MP-0 large sensitivity runs use the large checkpoint, CUDA execution, float64 precision, and dispersion enabled. The corresponding frozen settings are archived in the experiment configuration files under \texttt{research/agent\_eval/configs/}.

The four LLM backends evaluated in this study are: Gemini~2.5~Pro (\texttt{gemini-2.5-pro})~\cite{gemini2026modelcard}, Grok-4 (\texttt{grok-4-0709})~\cite{grok2025blog}, GPT-5.4 (\texttt{gpt-5.4-2026-03-05})~\cite{openai2026gpt5}, and Claude~Sonnet~4.6 (\texttt{claude-sonnet-4-6})~\cite{anthropic2026claude}. Access to these models requires API keys from their respective providers.

The CMU20 and OCD62 benchmark manifests are described in the main text. The CMU20 dataset is based on the Adsorb-Agent benchmark~\cite{ock2024adsorbagent}, while OCD62 is introduced in this work. These manifests are included in the GitHub repository under \texttt{datasets/}.
