\subsection{Controlled ablation on a locked benchmark set}

We evaluated the agent on a locked 15-case CMU benchmark subset (cases 01--02, 04--05, 09--10, 12--20), spanning atomic H, NNH, OH, acetaldehyde, and molecular CH$_3$ adsorption across increasingly difficult intermetallic and monometallic surfaces.
For each case, we executed five runtime variants under a fixed MACE-small CPU backend: \textit{Full}, \textit{$-$Slip}, \textit{$-$Forbid}, \textit{$-$Term}, and \textit{1-Shot}.
Here \textit{$-$Slip} removes slip-conditioned guidance, \textit{$-$Forbid} keeps slip detection but removes the explicit ``do not retry this site'' instruction, \textit{$-$Term} disables convergence-based early stopping, and \textit{1-Shot} reduces the total budget to a single attempt.
The full 15$\times$5 matrix was executed independently on four LLM backends---Gemini~2.5~Pro, Grok-4, GPT-5.4, and Claude~Sonnet~4.6---yielding 300 successful runs out of 300 (100\%).

\input{../research/results/paper_tables}

Tables~\ref{tab:si-ablation-gemini}--\ref{tab:si-ablation-claude} report the full energy matrices.
The Friedman test is highly significant on all four backends: Gemini ($p = 1.4 \times 10^{-6}$), Grok-4 ($p = 2.2 \times 10^{-6}$), GPT-5.4 ($p = 2.2 \times 10^{-3}$), and Claude ($p = 1.1 \times 10^{-4}$).
Pairwise Wilcoxon tests confirm that the full agent significantly outperforms the 1-shot baseline after BH correction on three of four backends (Gemini BH~$p = 0.006$, Grok-4 BH~$p = 0.011$, Claude BH~$p = 0.013$); GPT-5.4 is borderline (BH~$p = 0.084$).

\paragraph{The closed-loop gain is large and backend-dependent.}
The mean full-versus-single-shot improvement varies from 0.097~eV (GPT-5.4) to 0.363~eV (Gemini), with Grok-4 at 0.341~eV and Claude at 0.185~eV.
The largest per-case improvements occur on complex surfaces: case~20 shows 1-shot deficits exceeding 1.0~eV on both Gemini and Grok-4, while case~12 and case~15 consistently degrade under single-shot across all backends.
GPT-5.4 shows the smallest mean improvement, consistent with a stronger one-shot planner that leaves less room for iterative recovery; however, it still gains 0.426~eV on case~02 and 0.549~eV on case~17.

\paragraph{Ablated variants sometimes outperform full.}
On 15 cases, the expanded matrix reveals a systematic pattern: ablated variants find lower energies than the full agent on specific case$\times$backend combinations.
Case~18 is the clearest example: \textit{$-$Slip} improves over full by 0.491~eV on GPT-5.4 and Claude, while \textit{$-$Forbid} improves by 0.350~eV on Gemini.
Case~16 shows a reversal where \textit{$-$Slip} degrades by 0.391~eV on Claude but \textit{$-$Forbid}, \textit{$-$Term}, and 1-Shot all match or beat the full variant on GPT-5.4.
Case~17 similarly shows \textit{$-$Slip} outperforming full on Grok-4 by 0.280~eV.
These are not anomalies but a predictable consequence of the guardrail architecture: on surfaces where early attempts happen to find good configurations, the \textit{FORBID} or slip-conditioned constraints can over-prune the remaining search space, steering the agent away from solutions it would otherwise find.
The GPT-5.4 statistics make this concrete: \textit{$-$Forbid} and \textit{$-$Term} both trend toward improving over full (BH~$p = 0.084$, negative rank-biserial $r$), confirming that on this backend the guardrails impose a small net cost on the expanded case set.

\subsection{Termination mostly controls compute, not chemistry}

The termination signal behaves primarily as a cost-control mechanism across all four backends.
For Grok-4, the mean energy delta between \textit{Full} and \textit{$-$Term} is $-$0.012~eV; disabling termination matches or improves the full agent on 14 of 15 cases.
For Gemini, the same ablation has a mean delta of +0.018~eV, with the largest degradation on case~19 (+0.247~eV).
For GPT-5.4, the effect is marginally beneficial: mean delta $-$0.078~eV, driven by cases~12 and~18 where disabling termination allows the agent to find better solutions.
For Claude, the mean delta is $-$0.031~eV, similarly dominated by case~18 (+0.491~eV improvement).
However, the compute overhead of disabling termination remains substantial: token consumption can increase 2--6$\times$ on individual cases.
Thus, the termination logic should be understood as a guard against planner churn after the search has converged---removing it rarely helps and sometimes hurts, but always costs more tokens.

\subsection{Closed-loop search reduces backend variance}

Table~\ref{tab:cross-llm-summary} summarizes energy agreement across all four backends on the 75 case$\times$variant configurations (15~cases~$\times$~5~variants).
Under the iterative full agent, the mean four-backend range is 0.129~eV, with 7/15~cases agreeing within 0.01~eV.
Under the 1-shot baseline, the mean range rises to 0.425~eV with 0/15~within 0.01~eV---a 3.3$\times$ increase in backend spread.

The tightest convergence is achieved by the \textit{$-$Term} variant (mean range 0.077~eV, 9/15 within 0.01~eV), consistent with the observation that disabling termination allows all backends to converge to similar local minima given sufficient iterations.
The largest four-backend spread occurs in 1-shot case~20 (1.050~eV), where Gemini and Grok-4 find an energy basin near $-$10.6~eV while GPT-5.4 and Claude converge to $-$11.65~eV.
Under full iterative search, the same case narrows to a range of 0.001~eV---all four backends converge to effectively the same solution.
These data support a concrete systems claim: the closed-loop architecture acts as a backend-robustness mechanism that drives heterogeneous LLM planners toward the same physical solution.

\subsection{Independent validation on the OCD-GMAE dataset}

To test whether the ablation findings generalise beyond the 15 locked CMU cases, we constructed an independent validation set from the Open Catalyst 2020 Dense dataset~\cite{lan2023adsorbml}.
We selected 10 cases spanning 8 adsorbate species (H, OH, NH$_2$, N$_2$, NH, N, NO, N$_2$O), 9 distinct intermetallic surfaces, and a wide range of DFT reference energies ($-$14.55 to $-$1.35~eV).
The same 5$\times$5 ablation matrix was executed successfully on all four LLM backends (Gemini~2.5~Pro, Grok-4, GPT-5.4, Claude~Sonnet~4.6), totaling 200/200 runs. Gemini was accessed via Google Vertex AI direct API.

\input{../research/results/ocd_gmae_paper_tables}

\paragraph{Closed-loop improvement is larger and more consistent.}
The mean full-versus-single-shot improvement across the 10 cases is 0.542~eV (4-backend average), substantially larger than the 0.247~eV observed on the CMU benchmark.
Nine of ten cases show improvement exceeding 0.05~eV; only case~015 is effectively neutral (mean $\Delta = -0.011$~eV across backends).
The largest gains occur on cases with multiple competitive binding sites: case~004 (+1.325~eV on Gemini), case~013 (+1.284~eV on Gemini), and case~023 (+0.985~eV on Gemini).

\paragraph{Statistical significance is achieved.}
With 10 cases, the Friedman test reaches significance on all four backends: Gemini ($p = 3.4 \times 10^{-4}$), Grok-4 ($p = 0.002$), GPT-5.4 ($p = 0.007$), Claude ($p < 0.001$).
For GPT-5.4, the pairwise Wilcoxon test for $-$Slip versus Full is individually significant ($p = 0.016$), confirming that slip-conditioned feedback contributes measurably to search quality on this dataset.

\paragraph{Backend convergence pattern.}
Table~\ref{tab:ocd-cross-llm} reports the four-backend cross-LLM robustness analysis (Gemini accessed via Vertex AI direct API, the remaining three via their native endpoints).
The $-$Term variant achieves the tightest cross-backend agreement (mean range 0.240~eV), while single-shot shows the widest spread (0.810~eV)---a 3.4$\times$ ratio.
The overall pattern mirrors the CMU result but with higher absolute ranges, reflecting the greater chemical diversity and more complex potential-energy surfaces in the OCD-GMAE set.
The Full variant shows a mean range of 0.352~eV, driven largely by case~013, where Claude finds a distinct local minimum ($-$9.320~eV) that the other three backends miss ($-$10.361~eV).

\paragraph{Ablated variants can outperform full.}
Several cases show ablated variants finding lower energies than Full: case~004 under Grok-4 $-$Term ($-$11.384 vs.\ $-$10.138~eV, $\Delta = -1.247$~eV), case~005 under Grok-4 $-$Slip ($-$10.643 vs.\ $-$10.118~eV), and case~024 under Claude $-$Slip ($-$3.716 vs.\ $-$3.418~eV).
These confirm the CMU observation that the agentic guardrails are not universally beneficial: on certain surfaces, the FORBID or slip-conditioned constraints over-prune the search space.

\subsection{Comparison with Adsorb-Agent under identical physics}

To isolate the effect of search strategy from the choice of force field, we forked the open-source Adsorb-Agent codebase~\cite{adsorbagent} and replaced its EquiformerV2 relaxation backend with the same MACE-MP-0 (small, CPU, float32) calculator used throughout this work.
All remaining protocol parameters were locked to AdsMind's settings: BFGS optimizer, $f_\mathrm{max} = 0.10$~eV/\AA, 200 steps, and identical \textsc{FixAtoms} constraints (bottom third of slab atoms fixed).
We ran the modified Adsorb-Agent with GPT-4o (its default LLM) on all 20 benchmark cases.

\paragraph{Reliability gap.}
AdsMind succeeds on all 15 overlapping cases across every LLM backend (100\%).
Adsorb-Agent fails on 4 of these 15 cases (01, 09, 15, 20), producing zero candidate configurations because the LLM-predicted binding site type did not match any geometrically enumerable site on the surface.
This 27\% failure rate is a structural limitation of the single-pass architecture: if the initial site prediction is wrong, the system has no recovery mechanism, whereas AdsMind's closed-loop slip detection and FORBID constraints allow the agent to redirect after failed attempts.

\paragraph{Energy depth under matched physics.}
Among the 11 cases where both systems succeed, Adsorb-Agent finds a lower adsorption energy on 9, AdsMind wins on 1, and 1 is a tie ($|\Delta| < 0.01$~eV).
The mean energy difference is $+0.65$~eV in favor of Adsorb-Agent (Wilcoxon $p = 0.010$, $n = 11$).
The largest gains occur on complex adsorbates with many competitive binding sites: case~16 ($\Delta = 2.04$~eV), case~18 ($\Delta = 1.35$~eV), and case~17 ($\Delta = 1.17$~eV).
On simple adsorbates where the configuration space is small, the two systems converge to nearly identical energies (case~05, $\Delta = 0.008$~eV).

\paragraph{The gap is explained by sampling volume, not planning quality.}
Adsorb-Agent relaxes 6--41 initial configurations per case (median~21), exhaustively evaluating every site that matches the LLM's prediction.
AdsMind relaxes one configuration per iteration over $\sim$4 iterations, for a total of $\sim$4 relaxations per run.
The $\sim$5$\times$ compute-per-case difference directly explains why Adsorb-Agent reaches deeper energy basins when it does not fail: on surfaces with many local minima, more samples find better minima.
This is confirmed by the pattern of energy differences: the gap is negligible on simple surfaces and grows monotonically with adsorbate complexity (i.e., with configuration-space size).

\paragraph{Efficiency-normalized comparison.}
Per MACE relaxation, AdsMind's closed-loop gain over its own single-shot baseline (0.25--0.36~eV across backends; Section~3.1) is achieved in $\sim$4 relaxations, whereas Adsorb-Agent requires $\sim$21 relaxations to obtain a 0.65~eV advantage over AdsMind.
In energy gain per unit compute, the two strategies are comparable ($\sim$0.06--0.09~eV per relaxation for AdsMind vs.\ $\sim$0.03~eV per relaxation for Adsorb-Agent's marginal gain), but AdsMind delivers this gain with perfect reliability.
A compute-matched bootstrap analysis confirms this: when Adsorb-Agent's trajectories are subsampled to 4 per case (matching AdsMind's relaxation budget), AdsMind wins 32\% of bootstrap samples overall, with case-level win rates ranging from 0\% (cases where Adsorb-Agent finds a fundamentally different basin) to 100\% (simple surfaces where both systems converge).
The gap narrows most on complex adsorbates---case~16 shifts from a 2.04~eV deficit to a near coin-flip (48\% AdsMind win rate)---confirming that sampling volume, not planning quality, drives the energy difference.

\paragraph{Cost structure.}
The two systems have inverted cost profiles.
Adsorb-Agent consumes $\sim$4{,}000 LLM tokens per case (single planning pass) but $\sim$22 MACE relaxations; AdsMind consumes 25{,}000--40{,}000 LLM tokens per case (iterative closed-loop reasoning across $\sim$4 rounds) but only $\sim$4 MACE relaxations.
At current API pricing, the LLM cost difference is modest (\$0.02 vs.\ \$0.05--0.15 per case), but the $5.6\times$ MACE compute gap is substantial for large-scale screening campaigns where relaxation dominates wall-clock time.

\paragraph{Implications.}
The comparison reveals a reliability--depth trade-off between the two architectures.
AdsMind's closed-loop design guarantees convergence on every case and every backend, but its iterative single-configuration refinement under-samples the configuration space on complex surfaces.
Adsorb-Agent's parallel multi-configuration approach explores more of the energy landscape when the LLM's initial prediction is correct, but fails catastrophically when it is not.
A multi-seed control experiment confirms that the depth gap is architectural, not resource-limited: running AdsMind with 5 independent random seeds on the three worst cases (16, 17, 18) yields 20 total relaxations---matching Adsorb-Agent's budget---yet the remaining energy gap exceeds 1.1~eV on all three cases.
The iterative refinement trajectory stays within the same energy basin regardless of seed, whereas Adsorb-Agent's site enumeration reaches structurally distinct basins.
These findings motivate a concrete architectural extension: coupling AdsMind's closed-loop error recovery with Adsorb-Agent's systematic site enumeration at the initial placement stage.

\subsection{Non-LLM and cross-backend control baselines}

To isolate the contribution of the LLM planner from that of the underlying MACE relaxation pipeline, we implemented two non-LLM controls using the same MACE-matched physics: a \textit{random placement} baseline (N=20 random adsorbate poses per case, each relaxed with the standard protocol) and a \textit{heuristic enumeration} baseline (all high-symmetry sites identified by Autoadsorbate, median 47 sites per case).
We additionally ran Adsorb-Agent with GPT-5.4 to enable a same-LLM comparison with AdsMind.

\paragraph{Random placement matches AdsMind Full on average when given more samples.}
Across all 15 CMU cases, the random baseline achieves a mean energy difference of $-0.28$~eV relative to AdsMind Full (best-across-backends); it wins 4 cases, loses 3, and ties on 8.
On the two most complex cases it wins decisively---case~16 ($-6.74$ vs.\ $-4.66$ eV, $\Delta = -2.08$~eV) and case~20 ($-13.42$ vs.\ $-11.65$ eV, $\Delta = -1.77$~eV).
Critically, the random baseline uses $5\times$ more MACE relaxations per case (20 vs.\ $\sim$4).
AdsMind's \textit{1-Shot} variant, which allots the same single MACE relaxation as a naive search would, never beats AdsMind Full and underperforms random on several cases.
This confirms that the LLM planner's single-pass placement is not inherently better than random; the value of the agent lies in \textit{iterative refinement}, not initial guess quality.

\paragraph{Heuristic enumeration beats AdsMind Full on 6 of 15 cases.}
Systematically relaxing every Autoadsorbate-enumerated site (25--98 sites per case, no LLM) wins on 6 cases, loses on 1, ties on 8.
On case~19 (Hf$_2$Zn$_6$ + OCHCH$_3$), heuristic finds $-4.70$~eV versus AdsMind Full's $-4.04$~eV---a 0.66~eV gap comparable to Adsorb-Agent's.
The heuristic baseline uses 5--25$\times$ more MACE compute than AdsMind Full.
This result positions AdsMind's LLM planner as a \textit{compute-efficient approximation} of exhaustive site enumeration: AdsMind reaches within 0.19~eV of the heuristic optimum using 1/10 to 1/25 the number of relaxations.

\paragraph{Adsorb-Agent GPT-5.4 confirms the finding under a matched LLM.}
Running Adsorb-Agent with GPT-5.4 (the same backend used in AdsMind's strongest runs) produces a comparable outcome: 12 successful cases (vs.\ GPT-4o's 11 successes), with the same qualitative reliability--depth trade-off against AdsMind.
Failure cases differ between LLMs (GPT-4o fails on 01, 09, 15, 20; GPT-5.4 fails on 05, 15, 18), suggesting that single-pass LLM planning has an irreducible $\sim$20--27\% failure rate regardless of LLM capability (GPT-4o: 4/15~=~27\%; GPT-5.4: 3/15~=~20\%).
AdsMind's closed-loop architecture has a 0\% failure rate across all backends.

\paragraph{Iteration convergence.}
Tracing the per-iteration running-best energy across all 15 cases and 4 backends reveals that iteration 2 captures 58--73\% of the eventual full-run improvement over iteration~1, and iteration~3 captures 74--90\%.
The remaining iterations contribute diminishing returns.
This rapid convergence is the mechanistic basis for AdsMind's compute efficiency: most of the closed-loop value is extracted in the first 2--3 iterations.

\paragraph{MACE-large sensitivity.}
Repeating the full variant on 5 representative cases (05, 14, 16, 18, 20) with MACE-large (GPU, float64, dispersion) partially closes the gap to Adsorb-Agent on complex cases: case~16 AdsMind--Adsorb-Agent gap narrows from 2.04~eV (MACE-small) to 0.97~eV (MACE-large), and case~18 narrows from 1.35~eV to 0.39~eV.
The AdsMind--Adsorb-Agent energy ordering is preserved under both MACE variants, confirming that the conclusions drawn from MACE-small are qualitatively robust to force-field fidelity.
Case~20 failed under MACE-large due to dissociation detected at iteration~5.

\paragraph{Reframing AdsMind's contribution.}
Taken together, these controls reveal that AdsMind's primary contribution is not deepest-basin discovery.
Random sampling with 20 relaxations matches AdsMind Full on average; heuristic enumeration with $\sim$50 sites beats it on 40\% of cases; Adsorb-Agent with $\sim$21 relaxations and correct site prediction beats it on 60--80\% of cases.
AdsMind's distinctive value is the combination of: (i) 100\% success rate across backends and cases, versus 73--80\% for Adsorb-Agent; (ii) compute efficiency---matching random at $5\times$ less MACE work and approaching heuristic at $10$--$25\times$ less; (iii) chemical interpretability via Slip detection, FORBID logic, and backend-convergent reasoning traces absent from brute-force methods; and (iv) iterative error recovery, which is a prerequisite for autonomous chemistry workflows where silent failures are unacceptable.
A natural extension couples AdsMind's closed-loop error-recovery with broader initial sampling (multi-site LLM placement with parallel relaxation), which would combine the depth of enumeration with the reliability and interpretability of the closed-loop agent.
