\section{Discussion}

\subsection{What AdsMind contributes, and what it does not}

The results of this study, including the deliberately adversarial baselines introduced in Section~3, point to a nuanced characterization of closed-loop LLM agents that differs from the message conveyed by much of the prior literature.
Prior work on LLM agents for chemistry has tended to emphasize the \textit{capability} axis: what new kinds of tasks can an LLM accomplish that prior methods could not?
Our evidence suggests that for the specific problem of finding low-energy adsorbate configurations, the capability axis is not where closed-loop LLM reasoning wins.
Under identical MACE physics, random sampling with 20 relaxations matches AdsMind on eight of fifteen CMU cases and beats it on four; heuristic enumeration of $\sim$50 canonical sites beats AdsMind on six of fifteen; Adsorb-Agent's multi-configuration search beats AdsMind on 9--10 of 11--12 successful cases.
These are not artifacts of implementation---they hold across 4 independently trained LLM backends, across two independent benchmarks (CMU and OCD-GMAE), and across MACE-small and MACE-large force fields.

What closed-loop LLM reasoning contributes is along a different axis: the combined property of being \textit{reliable, compute-efficient, and interpretable} simultaneously.
Each of these attributes alone has precedents. Brute-force methods are reliable.
Heuristic enumeration is interpretable (canonical sites). Single-pass LLMs are compute-efficient.
What is new in AdsMind is the co-occurrence: 100\% success across every case and backend, 5--25$\times$ less MACE compute than the brute-force alternatives, and physically grounded reasoning artifacts (slip events, forbid constraints, site-mismatch fingerprints) that a practitioner can inspect, contest, and incorporate into downstream workflows.
This combination is a prerequisite for autonomous chemistry, where silent failure or irreproducibility in any one of these dimensions would break the pipeline.

\subsection{Chemical Slip as a standalone LLM evaluation signal}

Chemical Slip---the mismatch between an LLM-planned adsorption site and the site realized after physical relaxation---was introduced as a feedback signal for the closed-loop agent.
The slip statistics aggregated across 300 CMU benchmark runs and 200 OCD-GMAE runs reveal a secondary value: slip frequency, stratified by surface type and adsorbate, serves as a physically grounded proxy for LLM chemical reasoning quality that is independent of any downstream task metric.
The one-shot slip rate on intermetallic surfaces far exceeds the rate on monometallic surfaces across all four backends, confirming that LLM chemical priors are substantially less reliable on complex alloy compositions.
In contrast to text-based chemistry benchmarks (e.g., ChemBench) that test factual recall, Chemical Slip evaluates \textit{applied} reasoning---whether the model's structural prediction survives physical simulation---and complements existing LLM evaluation frameworks for scientific domains.

\subsection{Why brute-force sampling wins on depth, and when it matters}

The energy-depth advantage of random placement and heuristic enumeration on complex surfaces deserves explicit mechanistic explanation rather than hand-waving.
On simple adsorbates with few competitive binding sites (H on Cu$_3$Ag(111), OH on Pt(100)), AdsMind Full matches the non-LLM baselines within 0.01~eV; the PES has few distinct basins and iterative refinement from any initial guess converges to the global minimum.
On complex adsorbates (CH$_2$CH$_2$OH, OCHCH$_3$, ONN(CH$_3$)$_2$), the PES has many local minima separated by non-trivial barriers; iterative refinement from a single trajectory becomes trapped in whichever basin the first relaxation descends into, regardless of whether that is the global minimum.
Broad initial sampling circumvents this trap by seeding multiple trajectories in different basins.

This observation matters practically because it delineates when closed-loop search is the right tool.
For high-throughput catalyst screening where thousands of candidate surfaces must be down-selected, brute-force methods are more compute-efficient in aggregate (all relaxations run in parallel) and their failure modes are tolerable (failed cases can be re-queued).
For autonomous exploration of a single promising candidate where failure is costly and interpretability is required, closed-loop refinement adds value.
A two-stage pipeline---brute-force screening followed by closed-loop refinement on top candidates---would combine the throughput of the former with the reliability of the latter, and we outline the architectural integration in Section~4.3.

\subsection{MLFF as surrogate: scope and validation}

All experiments in this study use MACE-MP-0 small as the primary physical backend.
The agent's value is orthogonal to MLFF accuracy: it optimizes search strategy, not the energy function itself.
Inter-configuration energy differences on the same surface are primarily determined by local adsorbate--surface bonding geometry, which equivariant graph neural networks describe faithfully.
Systematic bias in absolute energies approximately cancels across configurations on the same surface and does not affect rankings.

To test this assumption directly, we repeated the AdsMind Full variant on 5 representative cases with MACE-large (GPU, float64, dispersion enabled).
The AdsMind--Adsorb-Agent energy ordering is preserved under both MACE variants on all 5 cases, and the cross-backend convergence pattern (Full < 1-Shot in backend spread) is preserved, confirming that the conclusions drawn from MACE-small are qualitatively robust to force-field fidelity.
The absolute magnitude of the AdsMind--Adsorb-Agent gap does narrow under MACE-large on complex cases (case~16: 2.04 $\to$ 0.97~eV; case~18: 1.35 $\to$ 0.39~eV), suggesting that some of the apparent gap under MACE-small reflects force-field artifacts rather than genuine search-strategy differences.
DFT validation of representative configurations is in progress through collaboration with the chemistry group.

Known MLFF failure modes---dispersion-dominated binding, strongly correlated $f$-electron systems, strong charge-transfer interfaces---define the current scope boundary of the framework.
Extending to these regimes will require either higher-fidelity MLFFs or hybrid MLFF--DFT workflows, but does not require changes to the agent architecture.

\subsection{Limitations and negative results}

\paragraph{Energy depth on complex adsorbates.}
On cases with multiple competitive binding sites and complex adsorbates (16, 17, 18, 19), AdsMind Full consistently produces higher energies than Adsorb-Agent, random, or heuristic baselines.
A multi-seed control experiment (5 independent random seeds on cases 16, 17, 18 with GPT-5.4) confirms this is architectural rather than resource-limited: the best-of-5-seeds AdsMind result remains more than 1.1~eV worse than Adsorb-Agent on all three cases.
The mechanism is the iterative refinement trajectory's tendency to stay within the initial energy basin regardless of seed.
We do not regard this as a defect of the architecture---it is inherent to any single-trajectory closed-loop refinement---but as a specification of the operating envelope in which closed-loop refinement adds value.

\paragraph{FORBID over-pruning.}
On several cases in both CMU and OCD-GMAE, removing the FORBID constraint \textit{improves} the final energy.
On GPT-5.4 the pairwise Wilcoxon test for $-$Forbid trends toward significance (BH $p = 0.084$) with negative rank-biserial correlation ($r = -0.79$), indicating that explicit negative constraints over-prune the search space on surfaces where the globally optimal site was initially rejected.
Adaptive or soft-constraint variants of the FORBID mechanism, in which the strength of a negative constraint decays with additional evidence, merit future investigation.

\paragraph{Statistical power.}
Although the 15-case CMU and 10-case OCD-GMAE benchmarks are large enough for Friedman tests to reach high significance across all four backends (min $p = 0.011$ on CMU, min $p = 3.4\times 10^{-4}$ on OCD-GMAE), many pairwise Wilcoxon tests do not survive BH correction.
We interpret individual pairwise comparisons primarily through effect sizes and bootstrap confidence intervals, and rely on the \textit{consistent direction} of effects across 4 independent LLM backends as a form of conceptual replication.

\paragraph{Scope.}
The current implementation handles flat and stepped crystalline surfaces with a single adsorbate species.
Extensions to defective surfaces, nanoparticles, co-adsorption, and solvated interfaces are architecturally straightforward but require additional validation.

\paragraph{LLM reproducibility and platform sensitivity.}
LLM APIs are black-box services subject to unannounced model updates; at temperature 0.0, outputs may still vary across API versions.
We mitigate this by recording exact model version identifiers, providing complete prompt templates (SI), and releasing full trajectory logs so that all reported results can be independently verified from saved data.
The four-backend design further ensures that conclusions do not depend on any single LLM provider.
A cross-machine control experiment on a subset of cases reveals that case~14 energies shift by $\sim$0.20~eV between Apple Silicon M3 Pro (macOS, ARM, float32) and EPFL x86\_64 (Linux, float32) platforms, while case~09 is stable within 0.003~eV; the shift is consistent across both Gemini and Grok-4 and preserves the within-run decision trajectory, indicating a platform/MLFF-environment effect rather than LLM divergence.
All within-backend ablation comparisons in the main text use a single platform throughout and are unaffected.

\section{Conclusion}

We have presented AdsMind, a closed-loop multi-agent framework that integrates LLM-driven planning with iterative MLFF relaxation feedback for autonomous adsorption configuration search.
Three mechanisms---Chemical Slip detection, FORBID constraint memory, and autonomous termination---enable the agent to detect its own reasoning errors, avoid repeating them, and stop when further exploration is unproductive.

Through 500 controlled runs across four LLM backends (Gemini, Grok-4, GPT-5.4, Claude) on two independent benchmarks (15-case CMU, 10-case OCD-GMAE), and head-to-head comparisons against four external baselines under identical MACE physics (random placement, heuristic enumeration, Adsorb-Agent with GPT-4o and GPT-5.4), we establish that AdsMind's primary value lies in the combination of (i) 100\% search success across every case and backend; (ii) cross-backend energy convergence (3.3$\times$ reduction in backend spread on CMU); (iii) compute efficiency (5--25$\times$ fewer MACE relaxations than brute-force controls); and (iv) interpretable reasoning artifacts in the form of slip events and forbid constraints.
We also report a substantive negative result: on complex adsorbates, brute-force sampling reaches lower energies than closed-loop refinement, a trade-off that arises from single-trajectory iteration rather than from insufficient compute, and one that motivates future hybrid architectures.
Chemical Slip, introduced here as a feedback signal, additionally serves as a physically grounded diagnostic for LLM chemical reasoning quality---a standalone methodological contribution that may extend beyond the application in this work.

The central message is that closed-loop LLM reasoning is not a replacement for brute-force configuration search on well-resourced problems; it is the foundation for a new class of \textit{autonomous chemistry workflows} where reliability, efficiency, and interpretability must be achieved simultaneously.
A natural next step is to combine AdsMind's closed-loop error recovery with broader initial site enumeration, yielding an agent that inherits the depth of brute-force methods and the reliability of closed-loop refinement.

Code, frozen experiment configurations, structured result files, all agent logs, baseline implementations, and comparison scripts are available at \url{https://github.com/AI4QC/AdsMind}.
